---
title: "Data Wrangling in R for Health Outcomes Research"
output: github_document
keep.md: true
---

The objective of this workshop is to provide an introduction to data wrangling in R using the tidyverse. By the end of this workshop, you should be able to:

1. Use a project oriented workflow and an R markdown document.
2. Perform simple data manipulations using the 5 dplyr verbs (`select`, `filter`, `arrange`, `mutate`, `summarise`, `group_by`), and chain these operations together using piping (`%>%`).
3. Join datasets together using the set of `join` functions.


## Objective 1: Setting up our workflow

First, let's load the [tidyverse](https://www.tidyverse.org/), which is an "an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures." The operate in parallel to, and often in place of, the set of packages included in base R.

```{r}
#install.packages("tidyverse")
library(tidyverse)

# some other packages that we'll need down the road
#install.packages("here")
library(here)
#install.packages("readr")
library(readr)
```


For this workshop, we're going to work with a dataset from the National Institute of Diabetes and Digestive and Kidney Diseases that I pulled from [kaggle](https://www.kaggle.com). The data was collected for the purpose of developing a model to predict the occurence of diabetes mellitus among a group of Pima Indians. You can find more details on the data [here](https://www.kaggle.com/uciml/pima-indians-diabetes-database).

Let's load the data into R and check it out:

```{r}
diabetes <- read_csv(here("Data/diabetes.csv"))
```


(Aside: Check out this [this](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/) blogpost by Jenny Bryan about best practices for getting your data into R. *Hint:* if you use code like this: 'rm(list=ls())' she will "set your computer on fire." 

## Explore the data

So what do we have here?

```{r}
glimpse(diabetes)
```

Let's get some basic summaries of each of the variables so that we have an idea of what's in there:

```{r}
summary(diabetes)
```

What are some things we might want to check when we first bring our data into R? Let's list them here:

  - ...


## Objective 2: Data manipulations using dplyr

(This summary is taken from the [STAT 545](http://stat545.com/) class notes.)

The six main dplyr functions for data manipulation are:

> - Pick variables by their names (`select()`).
> - Pick observations by their values (`filter()`).
> - Reorder the rows (`arrange()`).
> - Create new variables with functions of existing variables (`mutate()`).
> - Collapse many values down to a single summary (`summarise()`).

Then there's `group_by()`, which can be used in conjunction with all of these.

And `%>%` (piping) joins them all together into one happy family. 

---

### 1. `select()` works on variables

Let's get started with the first of our dplyr verbs, `select()`.

```{r}
select(diabetes, ID, Pregnancies, Glucose)
```

Or, equivalently:

```{r}
select(diabetes, ID:Glucose)
```

But be careful with not explicitly referencing variables, and *please* don't do this:

```{r}
select(diabetes, 1,2,3)
```

`select()` has a lot of handy tricks:

  - `everything()` to select all variables
  - `-` to drop variables. 
  - Use `=` within select to rename variables (*New_name = Old_name*)
  - Reposition variables based on their order in `select()`
  
I want to make Diabetes the first variable in my dataframe:

```{r}
select(diabetes, Diabetes, everything())
```


---

### 2. `filter()` works on rows

Select works on columns, and `filter()` works on rows.

Some useful logical expressions to refer back to: 

Logical expressions are governed by __relational operators__, that output either `TRUE` or `FALSE` based on the validity of the statement you give it. Here's a summary of the operators:

| Operation | Outputs `TRUE` or `FALSE` based on the validity of the statement... |
| ------ | ----- |
| `a == b` | `a` is equal to `b` |
| `a != b` | `a` is not equal to `b`. |
| `a > b` | `a` is greater than `b`. |
| `a < b` | `a` is less than `b`. |
| `a >= b` | `a` is greater than or equal to `b`. |
| `a <= b` | `a` is less than or equal to `b`. |
| `a %in% b` | `a` is an element in `b`. | 

(Full credit to [STAT 545](http://stat545.com/) for this handy table)


Let's start by filtering all observations in which blood glucose levels are greater than 100.

```{r}
filter(diabetes, Glucose>100)
```

How about only data for people who are Obsese?

```{r}
filter(diabetes, BMI=="Obese")
```

Now I want people who are obsese or morbidly obese...

```{r}
filter(diabetes, BMI=="Obese" | BMI=="Morbidly Obese")
```

**Exercises**

1. What's another way to select obese and morbidly obese people?


2. Filter observations for people who are 25 to 50 years old.


---

### 3. `arrange()` changes the order of rows

Okay, now that we've got `select` and `filter()` covered. Let's slip in `arrange()` quickly.

```{r}
arrange(diabetes, ID)
```

We can arrange by multiple variables. If you want a variable to be ordered the opposite way, just `desc()`!

Ordered by BMI, and then from oldest to youngest.

```{r}
arrange(diabetes, BMI, desc(Age))
```

How is arrange handling a categorical variable?


---

### 4. `%>%` chains operations together

Alrighty then, now that we've got a few tools, it's time to `%>%`!

The power of piping:

```{r}
diabetes %>%
      select(id=ID, Glucose, Diabetes) %>%
        filter(Glucose <= 100) %>%
          arrange(id)
```

Another example:

```{r}
diabetes %>% 
    filter(BMI=="Obese", Age>=50) %>% 
      select(ID, Diabetes) %>% 
        arrange(desc(ID))
```

What's happening here? Let's write it down:

  - ...


Now that we're familiar with `%>%`, let's keep using it as we learn our remaining two verbs: `mutate()` and `summarise()`


### Use `mutate()` to create new variables




### We can `summarise()` our data



### Challenge exercises

Let's stop here to give you a chance to play with your new tools. 

1. 


