---
title: "Data Wrangling in R for Health Outcomes Research"
output: html_document
---

The objective of this workshop is to provide an introduction to data wrangling in R using the tidyverse. By the end of this workshop, you should be able to:

1. Perform simple data manipulations using the 5 dplyr verbs (`select`, `filter`, `arrange`, `mutate`, `summarise`, `group_by`)
2. Chain these operations together using piping (`%>%`)
3. Join datasets together using the set of `join` functions


## Goal 1: Setting up our workflow

First, let's load the [tidyverse](https://www.tidyverse.org/), which is an "an opinionated collection of R packages designed for data science. All packages share an underlying design philosophy, grammar, and data structures."

```{r}
#install.packages("tidyverse")
library(tidyverse)
# some other packages that we'll need down the road
#install.packages("here")
library(here)
#install.packages("readr")
library(readr)
```


We're going to work with a dataset from the National Institute of Diabetes and Digestive and Kidney Diseases. The data was used to develop a model for predicting the occurence of diabetes mellitus using a variety of diagnostic measures. You can find more details on the data [here](https://www.kaggle.com/uciml/pima-indians-diabetes-database).

Let's load the data into R and check it out:

```{r}
diabetes <- read_csv(here("Data/diabetes.csv"))
```


(Aside: Check out this [this](https://www.tidyverse.org/articles/2017/12/workflow-vs-script/) blogpost by Jenny Bryan for best practices for getting your data into R. Hint: if you use this 'rm(list=ls())' she will "set your computer on fire." 


So what do we have here?

```{r}
str(diabetes)
```

Let's get some basic summaries of each of the variables so that we have an idea of what's in there:

```{r}
summary(diabetes)
```

What are some things we might want to know about this dataset? Let's list them here:

  - ...


## Topic 2: Simple data manipulations using dplyr

(This summary is taken from [STAT 545](http://stat545.com/) class notes.)

The six main dplyr functions for data manipulation are:

> - Pick variables by their names (`select()`).
> - Pick observations by their values (`filter()`).
> - Reorder the rows (`arrange()`).
> - Create new variables with functions of existing variables (`mutate()`).
> - Collapse many values down to a single summary (`summarise()`).

Then there's `group_by()`, which can be used in conjunction with all of these.

And `%>%` (piping) joins it all together into one happy family. 

---

Let's get started with the first of our dplyr verbs, `select()`.

```{r}
select(diabetes, ID, Pregnancies, Glucose)
```

Or, equivalently:

```{r}
select(diabetes, ID:Glucose)
```

But be careful with not explicitly referencing variables, and *please* don't do this:

```{r}
select(diabetes, 1,2,3)
```

---

Select works on columns, and `filter()` works on rows.

Some useful logical expressions to refer back to: 

Logical expressions are governed by __relational operators__, that output either `TRUE` or `FALSE` based on the validity of the statement you give it. Here's a summary of the operators:

| Operation | Outputs `TRUE` or `FALSE` based on the validity of the statement... |
| ------ | ----- |
| `a == b` | `a` is equal to `b` |
| `a != b` | `a` is not equal to `b`. |
| `a > b` | `a` is greater than `b`. |
| `a < b` | `a` is less than `b`. |
| `a >= b` | `a` is greater than or equal to `b`. |
| `a <= b` | `a` is less than or equal to `b`. |
| `a %in% b` | `a` is an element in `b`. | 

(Full creds to [STAT 545](http://stat545.com/) for this handy table)

Let's start by filtering all observations in which blood glucose levels are greater than 100.

```{r}
filter(diabetes, Glucose>100)
```

How about only data for people who are Obsese?

```{r}
filter(diabetes, BMI=="Obese")
```

Now I want people who are obsese or morbidly obese...

```{r}
filter(diabetes, BMI=="Obese" | BMI=="Morbidly Obese")
```

**Exercises**

1. What's another way to select obese and morbidly obese people?

2. Filter observations for people who are 25-50 years old.


---

Let's slip in `arrange()` quickly here.

```{r}
arrange(diabetes, ID)
```

Now I want to order by BMI, and then from oldest to youngest.

```{r}
arrange(diabetes, BMI, desc(Age))
```

How is arrange handling a categorical variable?


---

Alright then, now that we've covered `select()` `filter()` and `arrange()`  it's time to `%>%`!

The power of piping:

```{r}
diabetes %>%
      select(id=ID, Glucose, Diabetes) %>%
        filter(Glucose <= 100) %>%
          arrange(id)
```


One more example, and then ask what's happening here. 

Define a new variable with mutate.

Show a basic summarise. And then a group by and summarise - incorporate piping into all of these

Then define some larger challenge exercises using this dataset 



